National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs

### 2020-11-17 Analysis.Rmd
- include a source file to make Rmd manuscript more manageable - revised text

# Inference holds  FOCe INTERCEEpTS TO BE 
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
mod.2 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
summary(mod.2)

@@

Tables

@@

Plot models
```{r USEcoefficientplots}
plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_abyss() + scale_colour_brewer(labels = c("Personal Wellbeing", "Life Satisfaction")) # Plots
```

```{r cache = TRUE, dependson = "mod.1"}
#### Predicted Effects ####
pr.PWI <- ggpredict(model = mod.1, terms = c("Years [0:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.PWI, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Predicted Values of Personal Wellbeing") 
```
```{r pwi_expected}
```

```{r}
pr.LIFESAT <- ggpredict(model = mod.2, terms = c("Years [0:9]", "Beliefs"),
                        ci.lvl = 0.95,
                        type = "fe",
                        typical = "mean",
                     #   condition = c(Euro =1),
                        back.transform = TRUE,
                        ppd = FALSE,
                        interval = "confidence")
p2<-plot(pr.LIFESAT,facets = T) + 
 # scale_y_continuous(limits=c(6.35,6.85) ) + 
  gghighlight() + theme_blank() + 
  ggtitle("Predicted Values of Life Satisfaction") #+   scale_colour_brewer(palette = "Zissou1")
```

@@

### equations
Lifesatisfaction Model 1 equation

```{r}
library(equatiomatic)
print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3))
```

PWI Model 2 equation
```{r}
library(equatiomatic)
print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3))
```

@@

short report:
```{r, cache =TRUE, dependson=mod.1,include=FALSE}
report::model_text(mod.1)
```

@@

```{r, cache =TRUE, dependson=mod.1, include=FALSE}
r1<- report::report(mod.1)
report::text_long(r1)
```

### Model 2
For publication change all "significants" to "reliable"

short report
```{r, cache =TRUE, dependson=mod.2,include=FALSE}
report::model_text(mod.2)
```

>>>>

# Inference holds 
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
mod.2 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
summary(mod.2)

@@

@@

Plot models
```{r USEcoefficientplots}
plot_models(mod.2, mod.1) + ylim(-.4, .75) + theme_abyss() + scale_colour_brewer(labels = c("Personal Wellbeing", "Life Satisfaction")) # Plots
```
```{r}
library(parameters)
g1<-plot(parameters::model_parameters(mod.1))+ ggtitle("Personal Wellbeing Coefficients")
g2<-plot(parameters::model_parameters(mod.2))+ ggtitle("Life Satisfaction Coefficients")
```

```{r USE_2_coefficientplots}
g1/g2
```

```{r cache = TRUE, dependson = "mod.1"}
#### Predicted Effects ####
pr.PWI <- ggpredict(model = mod.1, terms = c("Years [7:9]","Beliefs"), 
                    ci.lvl = 0.95,
                    type = "fe",
                    typical = "mean",
                   # condition = c(Euro =1),
                    back.transform = TRUE,
                    ppd = FALSE,
                    interval = "confidence")
p1<-plot(pr.PWI, facets = T) + 
 # scale_y_continuous(limits=c(4.75,5.25) )  + 
  gghighlight()  +  theme_blank()+ 
  ggtitle("Predicted Values of Personal Wellbeing") 
```
```{r pwi_expected}
```

```{r predictlifesat}
pr.LIFESAT <- ggpredict(model = mod.2, terms = c("Years [0:9]", "Beliefs"),
                        ci.lvl = 0.95,
                        type = "fe",
                        typical = "mean",
                     #   condition = c(Euro =1),
                        back.transform = TRUE,
                        ppd = FALSE,
                        interval = "confidence")
p2<-plot(pr.LIFESAT,facets = T) + 
 # scale_y_continuous(limits=c(6.35,6.85) ) + 
  gghighlight() + theme_blank() + 
  ggtitle("Predicted Values of Life Satisfaction") #+   scale_colour_brewer(palette = "Zissou1")
p2
```

@@

### equations
Life satisfaction Model 1 equation

```{r eq1}
library(equatiomatic)
print(extract_eq(mod.1, wrap = TRUE,terms_per_line =3))
```

PWI Model 2 equation
```{r eq2}
library(equatiomatic)
print(extract_eq(mod.2, wrap = TRUE, use_coefs=TRUE,terms_per_line = 3))
```

@@

short report:
```{r, report1, cache =TRUE, dependson=mod.1,include=FALSE}
report::model_text(mod.1)
```

@@

```{r, cache =TRUE, dependson=mod.1, include=FALSE}
r1<- report::report(mod.1)
report::text_long(r1)
```
short report
```{r, cache =TRUE, dependson=mod.2,include=FALSE}
# Model 2
# For publication change all "significants" to "reliable"

report::model_text(mod.2)
```

### 2020-11-17 Manuscript.Rmd
- include a source file to make Rmd manuscript more manageable - revised text

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---

```{r include = FALSE}
library(knitr)
opts_chunk$set(
    fig.align = "center",
    root.dir = "../"
)
library("papaja")
library("tidyverse")
library("tinytex")
```

```{r load_libraries, echo=FALSE, include=FALSE}
# load libraries
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("sjPlot")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("gghighlight")
library("sjPlot")
library("see")
library("viridis")
library("ggpubr")
library("wesanderson")
library("ggsci")
library("here")
library("citr")
#library("furniture")
library("knitr")
library("texreg")
```
```{r load_data, include=FALSE}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```
```{r data processing, cache = TRUE, dependson = "ldf.5", include=FALSE}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

table1::label(dfben$Deprivation.S) <- "Deprivation (S)"
table1::label(dfben$Pol.Orient) <- "Political Conservative"
table1::label(dfben$Pol.Orient.S) <- "Political Conservative (S)"
table1::label(dfben$NZdep) <- "NZ Deprivation Index"
dfben$Partner <- factor(dfben$Partner, labels = c("No Partner","Has Partner"))
dfben$Employed <- factor(dfben$Employed, labels = c("Not Employed","Employed"))
table1::label(dfben$Edu.S) <- "Education (S)"

# create beliefs indicator
dfben$Beliefs<- factor(ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Not Believe Spirit", "_Skeptic_", 
                                        ifelse(dfben$Believe.God == "Not Believe God" & dfben$Believe.Spirit == "Believe Spirit", "_SpiritExcludesGod_",
                                               ifelse(dfben$Believe.God == "Believe God" & dfben$Believe.Spirit == "Believe Spirit","GodAndSpirit","GodExcludesSpirit"))))
```

The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

**Life Satisfaction** Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur]. Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

***Belief in a Spirit or a God***  We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

**Age** Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Education**  Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Deprivation** We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

**Employed** Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Partner** Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

**Ethnicity** Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

**Urban** People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

**Political Orientation** To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

**Equation 1**

@@

```{r, cache =TRUE, include =FALSE}
# model 2
mod.1 <-lmer (PWI ~  1 +  Years * Beliefs +  Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
```
```{r, cache=TRUE, include=FALSE}
# model 1
mod.1 <-lmer (LIFESAT ~ 1 + Years * Beliefs+ Age.10yrs.C + Deprivation.S + Edu.S  + Employed + EthnicCats + Male  + Partner  + Pol.Orient.S + Urban + (1|Id), data= dfben, na.action = na.omit)
summary(mod.1)
```

Results for both the personal well-being and life satisfaction models are presented in Table 2 and Figure 1.

Figure 1 about here:

@@

```{r, echo = FALSE, include=FALSE}
te00 <-texreg::extract(
  mod.1,
  level = 0.95,
  include.random = TRUE,
  include.rsquared = T,
  include.nobs = T,
  include.loo.ic = T,
  include.waic = F)
texreg(list(te00),
                custom.model.names = c("PWI"),
                caption = "Personal Wellbeing",
                sideways = F,
                scalebox = .5,
                #fontsize= "footnotesize",
                label = "tab:REGRESS_1",
                ci.force.level = 0.95, bold = 0.05,
                settingstars = 0,
                booktabs = TRUE,
                custom.note ="")
```

@@

**Interactions**

@@

**Interactions**

@@

<!-- ```{r} -->
<!-- knitr::include_graphics("Figs/Usecoefficientplots-1.png") -->
<!-- ``` -->

@@

@@

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples. Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.  We find that

>>>>

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")
```

```{r runanalysis, include = FALSE}
source('short.R')
```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

### Life Satisfaction
Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

### Personal well-being 
Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur]. Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

### Belief in a Spirit or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or life force" and "Do you believe in a god?" Responses were coded as (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who believe in a Spirit or Life Force; (3) Believers in a God who disbelieve in a Spirit or Life force; (4) Believers in a God and Spirit or Life Force. These believe were developed for the NZAVS from the 2005 Eurobarometer. For detailed information pertaining to beliefs, see Table 1.

@@

### Age 
Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Education
Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Deprivation
We measured the socio-economic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Employed
Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Partner
Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Ethnicity
Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Urban
People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Political Orientation
To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

## Equation 1

@@

@@

@@

##  Personal well-being: years X beliefs

@@

## Life Satisfaction: years X beliefs

@@

@@

```{r fig.cap = "Imputed data analysis: coefficient plot"}
knitr::include_graphics("Figs/coefficient_plot_both_imputed-1.png")
```

@@

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("Figs/predicted_PWI_impute-1.png")
```

Figure 5 about here

```{r fig.cap = "Imputed data analysis: predicted personal wellbeing by years X beliefs"}
knitr::include_graphics("Figs/predicted_lifesat_imputed-1.png")
```

@@

The purpose of this study has been to clarify the relationship between belief in a spirit/life force and subjective well-being, addressing three challenges from previous research: (1) tautological measures of spirituality and subjective well-being; (2) insufficiently distinct measures of spirituality and religion; (3) cross-sectional samples. Here, we ask whether beliefs in a Spirit or Life Force predict growth in personal well-being and life-satisfaction over eight years in populations who differ in the this belief, and who also differ in their belief in a God.

We find that \dots
1. Lower subjective well-being among spirit beliefs population at outset of study, surpassed (LS) or surpassing (PWI) by the end of the interval
2. Holds after adjusting for demograph
3. Holds for Belief in God

Questions:
1. Why initially lower well-being in NZ? 
2. Why the slope for spirit beliefs (and not God)?

Importance:
1. Methods: Disentangling
2. Growing population: practical importance
